Report on Late Winter2013 Production: Image Differencing

### 2013-03-22 W13report.tex
Names to subsections.

The focus of this production is to optimize and quantify the false
positive rate in the Psf--matched subtraction of a science image and a
reference image (referred to as the template image below).  The metric
we optimize is the number of false positives that are detected and
measured on the image difference.  We calculate several additional
quality metrics during the image subtraction process, and correlate
them with the number of false positives, as a means to predict the
number of false positives we may expect downstream from the image
subtraction.

\subsection{Algorithm and Background}

@@

\subsection{Implementation}

@@

All {\tt DiaSources} are associated with the {\tt calexp's} {\tt
  sources} and with the reference catalog, with a permissive matching
radius of 2''.  {\tt DiaSources} that match with {\tt sources} may be
true residuals around stars {\it or} true false positives that are
also found in {\tt calexp} SourceDetectionTask.  For this reason, we
use the associations with the reference catalog objects as the
assessment if a given {\tt DiaSource} matches with a {\tt source} or
whether it is an orphan (e.g. a statistical fluctuation in the
background).  In addition, all {\tt KernelCandidates} are persisted
with the above {\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories
are accessed with a specially written RepositoryIterator that
aggregates {\tt KernelCandidates} from a given {\tt source} across
multiple output repositories.

\subsubsection{Configs}

We ran several pre--productions to tune: the dimensions of the kernels
and KernelCandidates; the widths of the Gaussians (sigGauss) based on
the FWHM of the input images; and the scaling of the Gaussians
($\beta$) in the basis sequence.  The main production runs included:
an outer loop over the use of pre--filtering or post--filtering for
detection; variations of the complexity of the Kernel basis, including
the number of Gaussians (nGauss = 1,3) and the degree of the modifying
Laguerre polynomials (degGauss = 1..6); and the spatial order of the
Chebyshev polynomials on $a_i(x,y)$ (spatialKernelOrder = 1..6).
Finally, we ran several post--productions focused on perturbing the
solutions that yielded the fewest numbers of DiaSources along specific
dimensions.  This included: the RMS of the WCS solution used by
RegisterTask; including a DC offset in the source positions prior to
RegisterTask, to determine the ability of the kernel basis to
accommodate astrometric misregistration; the size of the SpatialCells
used in the spatial interpolation, which effectively regulates the
numbers of constraints that go into the spatial fit; and modifying the
detection thresholds to determine how the number of false positives
scales with threshold.

\subsection{Simulations}

@@

\subsection{Mean Squared Error}

@@

\subsection{Reduced $\chi^2$}

\section{False Positive Rate}

@@

A set of ``heat maps'' are presented in Figure~\ref{fp_heatmap}, that
demonstrate the total number of detections across all 9 sensors as a
function of the config parameters spatialKernelOrder and degGauss, for
nGauss = 3.  Qualitatively, these heat maps indicate that a Chebyshev
spatial order greater than 3 is required for all data.  In addition,
with the exception of the deconvolution case (visit 6866601 and
prefilter=False), the Gaussians need to be modified by Laguerre
polynomials of order 4 or higher.  The minimum of these detection maps
are found using the configs of highest complexity, with a very flat
slope in the number of false positives beyond spatialKernelOrder$>$3
and degGauss$>$3.  There is not much evidence for an upturn in the
numbers of false positives, meaning we have likely not reached the
case of overfitting using these configs.

@@

We have examined the distribution of these false detectiosn across the
image, and find no significant overdensities near the borders of the
images, which would be expected in cases of under or over--fitting of
the various spatial models (astrometry, backgrounds, Psf--matching
kernels).  Figure~\ref{edgedist} provides the distribution of
distances from the edge of the sensor for all false detections in {\it
  blue}, with a random sample of points in the image given in {\it
  green}.  Uncertainties are plotted using the square root of the
number of points in each bin, and the histograms are normalized to
provide a probability density.  The takeaway is that we find no
significant correlation of false positives with the boundaries of the
fitting functions.

\section{False Positive Dependencies}

After finding the configurations that yielded the minimal numbers of
false positives, we then perturbed the solutions to examine how the
numbers of false positives scaled with different effects.  This
included, in order of importance for this production: the difference
image detection threshold; the number of KernelCandidates going into
the spatial model; and astrometric registration errors.  We also
undertake a theoretical analysis of the number of false positives
expected in a 4k x 4k image as a function fo detection threshold, and
examine the requirements this sets on our ability to model the
variance and backgrounds in the images.  We compare the emperical
variation of false detections with detection threshold, and show that
our data are consistent with this.

\subsection{Detection Threshold}

\subsection{Number of KernelCandidates}

\subsection{Registration Errors}

@@

\subsubsection{Coordinate RMS}

@@

The false positive rate was not seen to vary significantly.

\subsubsection{Coordinate Offsets}

@@

\subsection{Background Subtraction}

\subsection{Number of False Positives With Detection Threshold}

To estimate the numbers of false positives we would expect in a
$4000\times4072$ pixel science image with a given FWHM in pixels, we
do the following.  The theoretical expectations for the number of
false positives due to random Gaussian fluctuations is outlined in
\cite{Kaiser-PointSources}.

\subsection{Variance Misestimation}

@@

\subsection{Measurements on Pre--filtered Data}

@@

We can reach near the theoretical limit for false positive rate in
these ideal data, something which has not been demonstrated for image
subtraction algorithms in the past.

Pre--filtering is clearly preferred when the traditional process would
lead to deconvolution, or sharpening, of the data.  When the standard
processing results in a convolution, or smoothing of the data,
pre--filtering provides fewer false positives.  As a trade--off, the
measurement of these false positives is not yet well defined.

@@

We intend to research additional aspects of the problem going forward,
including:
\begin{itemize}
\item How the false positive rate varies with more realistic SEDs;
\item How the false positive rate varies with airmass;
\item How the system behaves with a realistic mix of stars and galaxies;
\item How the false positive rate looks in crowded fields;
\item How to perform measurement on pre--filtered difference images;
\end{itemize}

@@

\clearpage
\begin{table}
\centering
\begin{tabular}{clcc}
\hline
\multicolumn{4}{|c|}{Best Results: Production 8} \\
\hline
Visit    & Prefilter & False Positive Matches & False Positive Orphans \\
\hline
v6866601 & True      & 5                      & 29  \\
         & False     & 10                     & 81  \\
v6866602 & True      & 5                      & 30  \\
         & False     & 9                      & 35  \\
v6866603 & True      & 2                      & 9   \\
         & False     & 1                      & 23  \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false positives.  Matches are determined using a
  2'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored. \label{tab-bestfp8}}
\end{table}

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{5}{|c|}{Best Results: Production 10} \\
\hline
Visit    & Prefilter & False Detections: Matches & False Detections: Orphans & Nneg / Npos \\
\hline
v6866601 & True      & 59                        & 70  & 1.8 \\ 
         & False     & 99                        & 44  & 2.9 \\
v6866602 & True      & 7                         & 35  & 2.2 \\
         & False     & 8                         & 45  & 0.7 \\
v6866603 & True      & 1                         & 22  & 2.8 \\
         & False     & 1                         & 35  & 3.0 \\
%v6866601 & True      & 11                        & 70  & 1.8 \\ SUBTRACT ME!
%         & False     & 99                        & 143 & 2.9 \\
%v6866602 & True      & 7                         & 42  & 2.2 \\
%         & False     & 8                         & 53  & 0.7 \\
%v6866603 & True      & 1                         & 23  & 2.8 \\
%         & False     & 1                         & 36  & 3.0 \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false positives.  Matches are determined using a
  2'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored.  The final column lists ratios of the
  number of negatively detected false detections to those with
  positive flux. \label{tab-bestfp10}}
\end{table}

@@

@@

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/cellsize.eps} \\
\caption{False positives with cell size
}
\label{cellsize}
\end{figure}

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/edge.eps} \\
\caption{Distribution of false positives as a function of the minimum
  distance from the edge of the sensor.  The {\it green} line shows
  the results from a random distribution of points.  The {\it blue}
  lines show the measured distributions, normalized as a probability
  density ($\sum y~\Delta x = 1$).  Error bars come from the square
  root of the number of points in each bin. }
\label{edgedist}
\end{figure}

>>>>

The focus of this production is to optimize and quantify the false
positive rate in the Psf--matched subtraction of a science image and a
reference image (referred to as the template image below).  The metric
we optimize is the number of false detections that are detected and
measured on the image difference.  We calculate several additional
quality metrics during the image subtraction process, and correlate
them with the number of false detections, as a means to predict the
number of false detections we may expect downstream from the image
subtraction.

\subsection{Algorithm and Background {\bf ANDYB}}

@@

\subsection{Implementation {\bf ANDYB}}

@@

All {\tt DiaSources} are associated with the {\tt calexp's} {\tt
  sources} and with the reference catalog, with a permissive matching
radius of 2''.  {\tt DiaSources} that match with {\tt sources} may be
true residuals around stars {\it or} true false detections that are
also found in {\tt calexp} SourceDetectionTask.  For this reason, we
use the associations with the reference catalog objects as the
assessment if a given {\tt DiaSource} matches with a {\tt source} or
whether it is an orphan (e.g. a statistical fluctuation in the
background).  In addition, all {\tt KernelCandidates} are persisted
with the above {\tt LOCAL} and {\tt SPATIAL} metrics.  Repositories
are accessed with a specially written RepositoryIterator that
aggregates {\tt KernelCandidates} from a given {\tt source} across
multiple output repositories.

\subsubsection{Configs {\bf ANDYB}}

We ran several pre--productions to tune: the dimensions of the kernels
and KernelCandidates; the widths of the Gaussians (sigGauss) based on
the FWHM of the input images; and the scaling of the Gaussians
($\beta$) in the basis sequence.  The main production runs included:
an outer loop over the use of pre--filtering or post--filtering for
detection; variations of the complexity of the Kernel basis, including
the number of Gaussians (nGauss = 1,3) and the degree of the modifying
Laguerre polynomials (degGauss = 1..6); and the spatial order of the
Chebyshev polynomials on $a_i(x,y)$ (spatialKernelOrder = 1..6).
Finally, we ran several post--productions focused on perturbing the
solutions that yielded the fewest numbers of DiaSources along specific
dimensions.  This included: the RMS of the WCS solution used by
RegisterTask; including a DC offset in the source positions prior to
RegisterTask, to determine the ability of the kernel basis to
accommodate astrometric misregistration; the size of the SpatialCells
used in the spatial interpolation, which effectively regulates the
numbers of constraints that go into the spatial fit; and modifying the
detection thresholds to determine how the number of false detections
scales with threshold.

\subsubsection{Measurement {\bf RUSS}}

\subsection{Simulations {\bf ANDYC}}

@@

\subsection{Mean Squared Error {\bf SIMON}}

@@

\subsection{Reduced $\chi^2$ {\bf SIMON}}

\section{False Detection Rate}

\subsection{Theoretical {\bf ANDYC}}

\subsection{Empirical {\bf ANDYB}}

@@

A set of ``heat maps'' are presented in Figure~\ref{fp_heatmap}, that
demonstrate the total number of detections across all 9 sensors as a
function of the config parameters spatialKernelOrder and degGauss, for
nGauss = 3.  Qualitatively, these heat maps indicate that a Chebyshev
spatial order greater than 3 is required for all data.  In addition,
with the exception of the deconvolution case (visit 6866601 and
prefilter=False), the Gaussians need to be modified by Laguerre
polynomials of order 4 or higher.  The minimum of these detection maps
are found using the configs of highest complexity, with a very flat
slope in the number of false detections beyond spatialKernelOrder$>$3
and degGauss$>$3.  There is not much evidence for an upturn in the
numbers of false detections, meaning we have likely not reached the
case of overfitting using these configs.

@@

We have examined the distribution of these false detectiosn across the
image, and find no significant overdensities near the borders of the
images, which would be expected in cases of under or over--fitting of
the various spatial models (astrometry, backgrounds, Psf--matching
kernels).  Figure~\ref{edgedist} provides the distribution of
distances from the edge of the sensor for all false detections in {\it
  blue}, with a random sample of points in the image given in {\it
  green}.  Uncertainties are plotted using the square root of the
number of points in each bin, and the histograms are normalized to
provide a probability density.  The takeaway is that we find no
significant correlation of false detetions with the boundaries of the
fitting functions.

\section{False Detecion Dependencies}

After finding the configurations that yielded the minimal numbers of
false detections, we then perturbed the solutions to examine how the
numbers of false detections scaled with different effects.  This
included, in order of importance for this production: the difference
image detection threshold; the number of KernelCandidates going into
the spatial model; and astrometric registration errors.  We also
undertake a theoretical analysis of the number of false detections
expected in a 4k x 4k image as a function fo detection threshold, and
examine the requirements this sets on our ability to model the
variance and backgrounds in the images.  We compare the emperical
variation of false detections with detection threshold, and show that
our data are consistent with this.

\subsection{Detection Threshold {\bf SIMON}}

To estimate the numbers of false detections we would expect in a
$4000\times4072$ pixel science image with a given FWHM in pixels, we
do the following.  The theoretical expectations for the number of
false detections due to random Gaussian fluctuations is outlined in
\cite{Kaiser-PointSources}.

\subsection{Number of KernelCandidates {\bf ANDYB}}

\subsection{Registration Errors {\bf ANDYB}}

@@

\subsubsection{Coordinate RMS {\bf ANDYB}}

@@

\subsubsection{Coordinate Offsets {\bf ANDYB}}

@@

\subsection{Background Subtraction {\bf SIMON}}

\subsection{Variance Misestimation {\bf ANDYB/SIMON}}

@@

@@

We can reach near the theoretical limit for false detection rate in
these ideal data, something which has not been demonstrated for image
subtraction algorithms in the past.

Pre--filtering is clearly preferred when the traditional process would
lead to deconvolution, or sharpening, of the data.  When the standard
processing results in a convolution, or smoothing of the data,
pre--filtering provides fewer false detections.  As a trade--off, the
measurement of these false detetions is not yet well defined.

@@

We intend to research additional aspects of the problem going forward,
including:
\begin{itemize}
\item How the false detection rate varies with more realistic SEDs;
\item How the false detection rate varies with airmass;
\item How the system behaves with a realistic mix of stars and galaxies;
\item How the false detection rate looks in crowded fields;
\item How to perform measurement on pre--filtered difference images;
\end{itemize}

@@

\clearpage
\begin{table}
\centering
\begin{tabular}{clcc}
\hline
\multicolumn{4}{|c|}{Best Results: Production 8} \\
\hline
Visit    & Prefilter & False Detection Matches & False Detection Orphans \\
\hline
v6866601 & True      & 5                      & 29  \\
         & False     & 10                     & 81  \\
v6866602 & True      & 5                      & 30  \\
         & False     & 9                      & 35  \\
v6866603 & True      & 2                      & 9   \\
         & False     & 1                      & 23  \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false detections.  Matches are determined using a
  2'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored. \label{tab-bestfp8}}
\end{table}

\begin{table}
\centering
\begin{tabular}{clccc}
\hline
\multicolumn{5}{|c|}{Best Results: Production 10} \\
\hline
Visit    & Prefilter & False Detections: Matches & False Detections: Orphans & Nneg / Npos \\
\hline
v6866601 & True      & 11                        & 59  & 1.8 \\ 
         & False     & 99                        & 44  & 2.9 \\
v6866602 & True      & 7                         & 35  & 2.2 \\
         & False     & 8                         & 45  & 0.7 \\
v6866603 & True      & 1                         & 22  & 2.8 \\
         & False     & 1                         & 35  & 3.0 \\
%v6866601 & True      & 11                        & 70  & 1.8 \\ SUBTRACT ME!
%         & False     & 99                        & 143 & 2.9 \\
%v6866602 & True      & 7                         & 42  & 2.2 \\
%         & False     & 8                         & 53  & 0.7 \\
%v6866603 & True      & 1                         & 23  & 2.8 \\
%         & False     & 1                         & 36  & 3.0 \\
\end{tabular}
\caption{Total number of {\tt DiaSources} detected at 5--sigma from
  all 9 sensors in raft 2,2, for the configurations that yielded the
  lowest number of false detections.  Matches are determined using a
  2'' match radius with the input reference catalog.  Objects with the
  {\tt EDGE} flag are ignored.  The final column lists ratios of the
  number of negatively detected false detections to those with
  positive flux. \label{tab-bestfp10}}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{clcc}
\hline
\multicolumn{4}{|c|}{Variance Ratios} \\
\hline
Visit    & Prefilter & Mean Ratio & RMS Ratio \\
\hline
v6866601 & True      & 1.009      & 0.002     \\
         & False     & 1.005      & 0.001     \\
v6866602 & True      & 1.003      & 0.0004    \\
         & False     & 1.005      & 0.0007    \\
v6866603 & True      & 1.004      & 0.0003    \\
         & False     & 1.008      & 0.002     \\
\end{tabular}
\caption{Ratio of the emperical RMS in the difference images,
  calculated via the interquartile range, to the square root of the
  median of the variance plane.  In all cases the variance plane
  represents an {\it underestimate} of the true variance in the
  images.  We report the mean and RMS across all sensors.
}
\end{table}

@@

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/cellsize.eps} \\
\caption{False detections with cell size
}
\label{cellsize}
\end{figure}

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/edge.eps} \\
\caption{Distribution of false detections as a function of the minimum
  distance from the edge of the sensor.  The {\it green} line shows
  the results from a random distribution of points.  The {\it blue}
  lines show the measured distributions, normalized as a probability
  density ($\sum y~\Delta x = 1$).  Error bars come from the square
  root of the number of points in each bin. }
\label{edgedist}
\end{figure}
